\chapter*{Questions and additions}\addcontentsline{toc}{chapter}{Questions and additions}\setcounter{PARAGRAPH}{0}\section*{Notes}\addcontentsline{toc}{section}{Notes/additions}\ad{Quietism}We saw what happened after Eckhart: Suso++... This kind of thinking may, indeed,invite to quietism and even when it does not invite, it may lead to it...\ad{distinction but no essence}We can imagine a \co{distinction} without \co{ipseity} when we encounter something for the very first time: we do not re-cognize it, we do not know, nor even feel, what this thing, this feeling, this something might be. But we know that it is {\em something}, only that we do not know what, we encounter it as something different from everything else.\ad{holiness, concepts}Before you meet a holy man, you may have a very confused, if any, concept of holiness. But it is no goal to have all the concepts correct and ready in advance. The goal is only this: {\em when} you meet a holy man, do not let your concepts (or misconceptions) prevent you from recognizing his holiness, from admitting it. (Actually, concepts, no matter what the hermeneutics tell, make it harder to accept the testimony of new experiences.)\ad{actuality confused with eternity}One of the main effects common to all the drugs is that they reduce the world to the pure \herenow, they narrow the horison of perception and consideration to this very moment, without tomorrow, without yonder, to pure actuality so seductively resembling the mystical presenceof eternity.\ad{Analytical philosophy}says \wo{this is bubbling, incomprehensible rubbish}. Well, if I read only one page from Kant, it is incomprehensible, but it helps if I read more and think about it. When I read Bible for the first time I understood much less than when I read it again. But so was it when I read Carnap, and when I studied mathematics!\ad{expression vs. expressed}The question about truth has been all the time confused with the question about the formulation of truth; the absolute with the expression of the absolute. And finding no unique, unequivocal, verifiable formulation, one decided to abandon the quest.There is a deep lack of respect, both for the truth and for the other in this. The truth need not -- {\em can not} -- be expressed in a single way like a mathematical definition. (It is abundant, overflowing, non-actual.) And the other can understand the intention beyond the partial, even misleading expressions.\ad{reflection: USED}The art of reflection is not to jump to a higher level of self-consciousness and observe everything, oneself included, from above and outside. On the contrary! It is to stay inside and from within to record the events, feelings passing by without externalizing them. It is to realize that there is nothing outside, that everything is and must be seen only from within. The art of reflection is not to oppose immediacy but to preserve it -- in spite of reflection.\ad{Community-personality}There is a bunch of thinkers who tend to consider \thi{the state of the  culture}, \thi{the state of the civilisation} and to design their philosophyas if in response to the observed crisis. Admirable as such pieces of sociology ormight be, they pass often for philosophy -- the morewarmly welcome and embraced, the more actualities they manage to addressexplicitly.The hero of this mixture of sociological analysis of the crises with thepretensions to human, personal relevance is Nietzsche -- the depths of hissocio- and meta-socio- observations is the more striking, the easier onerecognizes in them the element of down to earth psychology. Now, we do not wantto criticise the analyses of culture and civilisation rooted in the descriptionof the involved personal values and characters. But we would limit them topurely socio-, or meta-socio- dimension and delegate, perhaps, to some corner ofsocio-philosophy of politics.\ftnt{The \wo{post} of  modernism could be taken as an abbreviation for such a meta-socio- of a  reflection dissolving individual in the play of market forces (just like the  criticised capitalism does) or some other forces, or else for a socio-meta-, a  sociology of    intellectualism incapable of anything more than games of meta-perspectives   and meta-interpretations. \citf{As the horison of the century which will be     Your century, there appears increase of complexity in most areas, including   the area of various \wo{ways of life}, of daily life. And this dictates the   decisive challenge: make sure that humankind will be able to accommodate to   very complex means of perceiving, understanding and acting, which transcend   beyond its expectations.}{Lyotard, Le Postmoderne expliqu\'{e} aux     enfants, VIII [p.113]} How do you make sure anything like that? And what do I care if humankind is able to accommodate to anything? I have to accommodate to all kinds of things, not any humankind...}  The relevance of such thinking may be to tell an indivdual where he is, in what world he happens to live. So far so good. Butthen, trying to philosophise, one tends to formulate programs, advice,suggestions -- committing the crime of passing from \thi{being} to \thi{ought}which one used to be the first to denounce.\ad{pantheism vs. gnosticism}Pantheism (and we observed its relationship to empiricism) recognizing value andholines of everything turns their totality into god. Gnosticism is the opposite:only the highest, inaccessible, ineffable divinity has any value, while all thedetails of actual world and life are valueless, if not directly evil. This is,indeed, the tendency of Heidegger's complaining about the forgetfulness of Beingand, at the same time, claiming the essential association of such aforgetfulness with the openness of existence. This is also the tendencydiscernible in Jung, the tendency which Gilson wanted to discern in mostinfluences of neo-platonism... Do not we also commit the same mistake?Hopefully not. We have said: as below so above, works on earth do influence thedecisions in heaven; and although both ontological and eventual concretefounding proceed from above, the transition from the former to the latter, andthe latter's particular form, are helped -- though never determined! -- byeverything happeing here below. \section*{Questions}\addcontentsline{toc}{section}{Questions}%% {\large{\paCheck if, verify, and so explain \equin. A is \equi\ B:\begin{enumerate}\itemthen A can be explained using B, and B using A; or\itemthen A can not be explained without B, nor B without A; or\item \ldots ?\end{enumerate}Perhaps, \thi{explaining} is too strong here \ldots\pa One vs.  Many, Section \refsp{sub:OneMany} \ldots needs fixing: Iguess, on the one hand, as a \co{nexus} of \co{aspects} -- this is ok(the whole, static); but, on the other hand, I did describe it as an{\em emergence}, as a \thi{generation} -- this, probably, is but\co{virtuality} (but isn't \co{virtuality} merely a word, then?)\paAttention (non-reactive) vs. will (control) may need more precise elaboration\refpp{pa:SchelerZ}\paI guess, the \co{non-actual experience} of \co{non-actuals} (\co{I}, \co{world}, etc.) should be described more concretely \ldots as \co{experience} but not \co{an experience}\ldots\paMore specifically about \co{commands} and \co{inspirations} -- they may be felt, as \co{actual signs}, but what they reveal is completely contentless. That is, they may be slightly more specific \ldots So, is there {\em the} ultimate form of a \co{command} which is completely contentless? I guess, this coincides with \sch. \paElaborate more on \thi{upward} movements. 1) As in memory, i.e., things going into virtual state, or else 2) by mere applications of reflective categories to higher levels, as in \refp{pa:dissociateUp} (chap.II, level-4: my-Self vs. Self)Also, on construction of \co{complexes}, as in \refp{pa:lab}.f\tsep{questions?}\paOn the one hand: asymmetry of Being, and on the other: \co{One} has no precedence,appears only in the triunity of aspects\co{existence}-\co{confrontation}-\co{One}. \ad{evil -- alienation, or not ... vs. death}On the one hand: evil is \co{alienation}; then: death is not so bad, since it iseventual dissolution in the \co{indistinctness}. So why isn't killing good? (Notbecause death must be willed, which might then justify suicide.)Perhaps: evil, i.e., \co{alienation} is that which inhibits \co{existence} in\co{existing}, which prevents it from \co{confronting} its\co{origin} (\refpp{alienX}). While dissolution in \co{indistinct}, death,overcomes (possibility of) \co{alienation} only because it dissolves one ofthe terms of \co{confrontation} -- the \co{existence} -- in the other.It follows that death, as such, is not evil. It does not follow that it is good,either. \paAt the level of immediacy (e.g., \refpp{pa:levsSigns} and earlier), the sign coincides with the signified and yet, the object is given as transcendent -- external? (? \oss\ vs. \rss ?)\ad{subjectivism, projections, atheism or ...}Non-\co{incarnated} ideals are worthless -- haven't I made up such an ideal (of \co{spiritual choice} and \co{love})?Even more significantly: is not this whole ``explanation by projection''? One,God, invisibles, etc., all are but poles of relations of which other pole isexistence.This, perhaps, is the difference -- a pole of a relation is not aprojection, even if its necessary condition is the other pole. ?..Existence isrelational..?There is no subjectivism -- without existence, God would make no sense, but thisdoes not mean that God can be reduced to any actual dimension of existence likethinking, feeling or the like. None of human \thi{faculties}, especially none ofthose pertaining to the sphere of visibility, is ever capable of approaching,let alone representing God...Yet, it smells pure a-theism, Schoppenhauer if not Fichte, for this God is not\thi{there, in-itself}, but only \thi{for-us}... Well...\ad{spirituial choice vs. dualism}It does remind about dualisms, religions of radical choice --zoroastrianism, mitraism, manicheim and various gnostic trends with bogomilismand katars as the closest relatives. Yet it is not... Ontologically, there isOne, and only that (it is not the most real among the reals, but just thereality of everything); spiritually, \No\ is not of the same order, although atthe same level, for it does not reach the bottom, the origin\paIs not distinguishing smth. \thi{in the world} {\em very} different from distinguishing smth. \thi{within myself}? (This probably goes away\ldots)%% }} %\end \large