\NOWY{Why?}\plan{Why}\paA word is an actual sign which points to something invisible, lyingbeyond the horizon of actuality. And since it transcends this horizon,it can never be fully present within its limits. What you can do is totry to apply -- and more often invent -- words which point towardsthings you intend in the way in which you apprehend them; to create asystem of signs which convey your feeling of life. It will never beunique, the only possible, final. But if you manage to convey thisfeeling -- which unlike its expression is much more universal andcommon -- then what else would you like to do with your signs?\paReductionism is a project of scientific consciousness. Perhaps it istrue, perhaps everything we experience can be reduced to some basiclevel of physical or chemical reactions. But the point is -- so what?How can this ``wisdom'' affect my life and actions? Is it at allrelevant whether this is so or not? Suppose that somebody produces adefinite proof that it is so. And? Shall I change anything in my lifebecause of such a proof?The only way such an ideology might become relevant would be toproduce some controlling mechanisms.  Perhaps, having traced everythingto such a basic level will enable us to control everything which isabove -- and as some maintain, independent from -- it. Then it mightbe a different story. But so far, the very meagerness of such resultsmakes the whole project irrelevant. Let those who believe in it, workon it. In the meantime, let me tell you a different story.\paIf I announce some problem uninteresting it is not because it isillegitimate for some reasons, because it should not be asked, or should be asked within ontology and not within ethics. I reject it because I do not find it relevant for my thinking, because it lies beyond my world.All problems disappearing from the philosophical scene do so not because they have been solved but because people got tired of them -- they become irrelevant.\paWszystkie osoby i sytuacje przedstawione tutaj sa\; rzeczywiste, leczich ewentualna zbie\.zno\'s\'c z sytuacjami rzeczywistymi jest czysto przypadkowa.\NOWY{Philosophy}\plan{Philosophy}\pa\wo{Inventive}, when applied to a philosopher, means almost the same as\wo{petty} -- the only difference is the additional element of \wo{smartness},which is not contained in every \wo{pettiness}.\paPhilosophy does not create -- it reconstructs.\paPhilosophy is the art of appreciating simplifications.\paNothing new is ever said; there is no thought which has not beenthought before. What is new may be interesting for a scientist, for an intellectual --never for a human being. Novelty is an idol.  Typically, what is old is noble. But it is not noble because it isold. On the contrary, it is old because it is noble.  Otherwise, itwould die as so many fashions and transient distractions. Philosophyis a series of signatures put by different thinkers under thoughtsexpressed earlier by somebody else. And so on, indefinitely, back intothe individual childhood or common, mythical prehistory.  But thisrecurrence is its strength -- not a weakness against which it shouldbe defended.\com{Traditionally, one would say that philospohy does not deal with \wo{concrete} things but with their \co{essences}. It does not speak about\thi{father} or \thi{table} but about \co{fatherhood} and \co{table-ness}.This view is still biased towards concrete {\em things}. But in truth, it deals withsomething which is \wo{above} the things and this world. Only that this something isnot the essence of things -- it is a different order of the world. It is the order wherequite different categories play the significant role, and where concrete things of thisworld are merely their possible manifestations, incarnations. And, of course, in differentcontexts very different things and very different relations between things may expressthe same aspects of this \HH\ order.}\pa``Any good text has many meanings''. (Hemingway)Would not this be the difference between philosophy and literature.\paLife makes \HH\ present; philosophy makes it \LL.\pa A person is a way of living and feeling. A philosophy, any philosophy, is a way of speaking. ``Mouth speaks from the richness of the heart.''  \paIt isn't so, as Russell would like it, that philosophy, starting with some obvious truths, attempts to arrive at astonishing paradoxes. (This is an attitude of a scientist or an arrogant intellectual and it seems that Russell was both. A philosopher shouldn't be either.) On the contrary, philosophy attempts to arrive at the most trivial conclusions known to all people and, hopefully, to the philosopher himself. The challenge, and not the least one, lies in finding an appropriate set of symbols -- words -- which is capable of expressing these conclusions as lucidly and coherently as possible; in making the \HH, which is known intuitively, \LL.\paContradictions are inventions of the separating activity of reason.There are no two things in life which couldn't go together.\qpa{Each difference is only a difference of degree. It is only within the horizon of actuality, where things appear under distinct signs, that one can claim absolute, essential, definite, etc. difference between two things. But what one is actually separating in this way are not the things but their actual signs.}\upa{Say, roughly: epistemology inquires into the mechanisms of acquiring knowledge about what is there, ontology into what is there, and ethics into how we should relate to what is there. (It does not matter if ``what is there'' isn't really, but only appears as being there. The point is that the phrase has some meaning to us.)Now, an inquiry into what is there is a way of relating to it, so epistemology is a part of ethics. Then in order to relate to what is there, we have to have some access to it, so ethics is a part of ontology. And finally, we cannot speak about what is there without relating to it in an understanding way, so ontology is a part of epistemology. Well, I do not object to the existence of {\em purely} epistemological, etc. questions. They may be meaningful in their narrow sense. And, of course, one can produce all kinds of distinctions and specific methods distinguishing these disciplines. But nevertheless, from the very beginning, the division is unfounded if we are interested in ako. totality of our being. Separating these things and then trying to build bridges joining them together is a weird procedure:first one tries to emphasize the distinguishing features, and then to ignore them. I do not think that it is a fruitful point of departure for philosophical thinking which attempts to create a general picture of man's life and being in the world.}\pa\label{onethough}Each great philosophy expresses (at least) one deep intuition. Typically, it has become great by blowing up this single thought to the universal and only principle of everything.\paAn analytical philosopher is a guy who will ask you, when after a concert you say ``It was a beautiful piece of music'', ``What do you mean by `beautiful'?'', ``What do you mean by `a piece of music'?'' \\(It is the same usurpation as in \refp{onethough}.)\paAnalytical ``philosophers'' have to put almost all philosophy in the quotation marks. Well, I prefer to save some ink by putting only their ``philosophy'' into such marks.\paIn fact, they are no better than all these earlier fractions and thinkers whom they accuse of producing only unverifiable Weltanschaungen. They are much worse -- to the degree that their Weltanschaung does not concern any world but only a method. In their pretty out-of-date fascination with science they forbid themselves to talk about views and matters that matter, and are allowed to talk only about what philosophy should and should not be.\paYet, analytical philosophy is one of the most respectable forms of this great complex of the modern intellectuals: to get closer to people, to the``man on the street'', to the ``concrete, real world''.\paIt often strikes me that if we take any text on transcendental philosophy(whether Kant, Husserl, or whoever) and replace the word ``transcendetal''with ``unconcious'', we get equally meaningful yet much more comprehensibletext.\pa\thi{Brain in the box} -- silly. What does it show? That, perhpas, the wordis not the way the brain imagines it to be. And that's it! It does not showthat, perhaps, the word is not there, only that it may be different. For ithas to assume that the brain is connected to something outside itself. Andfuthermore, it does not show that the brain couldn't come up with the ideathat, perhaps, the word it `sees' is not the way it is, that, perhaps, ititself is only a brain in the box.\paAn academic attitude forces a philosopher to relate to and take into account any possible deisinction. A personal attitude is to stop distinguishing at some point.\NOWY{Words}\plan{Words}\paA text, the celebrated \thi{text}, is but a limited medium of communication. Limited by the absence of the author -- if the author could be present, the texts would rpobably be much shorter.Those who make \thi{text} into an independent, absolute event of cultural world forget or, as the matter may be, want to eliminate the fact that any text actually does or did have an author. Even if we know nothing about him except that he was the author of the \thi{text} we are actually reading, it is already more than enough. The \thi{text} is but a medium of communication -- if not from the author to the reader so, in any case, from one human being to another. And even if \wo{author's intentions} may be an unfortunate term, what we try to find in a \thi{text} can be so labeled -- it is some human meaning, something one human being might want to say to another.\paIn a way, everything is matter of speaking, of saying things. Words try to reflect our experiences but, equally, influence the way we see and act.\paThe world is but a sign.\paExpressions come before words; words before concepts. (Individual names before the common ones.)\paWord -- a sign -- is a prereflective synthesis which precedes any analytical understanding.\paA word whose meaning can be fully analyzed is redundant -- itfunctions merely as a convenient abstraction increasing efficiency of the system.\upa{\thi{Big words}, or better \thi{high words}, do not force their meaningupon us. They only hint at something not fully expressible, which weare free to model and interpret -- they leave us freedom, exactlybecause they do not have a unique, precise meaning.As such, they are the opposite of arguments which, in honesty orarrogance, always attempt to force the other to accept them.}\paWords never embrace the whole reality -- they are mere signs,pointers. If you do not understand what is being said, perhaps, you donot know what the talk is about. And if you know, you need not theabsolute precision of expression -- a mere sign will suffice.\paThe difference between \thi{words} and \thi{mere words} is exactly this: the latter fail to make anything present, while the former do reveal, that is, make the other realize what they are pointing to. \pa Communication means sharing the \wo{object} of discourse. With somebody we know very well, we can talk in very imprecise terms and still be certain that we are actually communicating. Often, because, when we have only begun to utter a sentence, the other already knows what we are trying to say. With strangers, we usually have to be much more precise.The art of speaking consists in such a \co{making the \wo{object} present and accessible} to others. It has nothing to do with the precision and definiteness -- these become necessary when communication is threatened.\pa``The primitive mind ascribes conscious life to all natural objects,to nature in general.'' The modern mind ascribes to it nothing at allor, perhaps, some crude functionalism. In either case, it is a matterof ascribing.\paTo some extend words are irrelevant. Manicheism is as good as Augustine'sChristianity. The difference concerns the implied attitude.But it is not at all the question of a subjective image of God. It isa question of being, of existential realization of faith, which hasinfinite variety of expressions. The words should merely point towardssuch expressions.\paA dialog is to look at the world also from another point of view.Dialog with oneself is no contradiction (rather with one's Self).\paAre there thoughts I did not think? Certainly. It doesn't botherme. What bothers me sometimes is that there are thoughts which I {\emcannot} think because there are things I have not seen.\paThe power of language consists in that we can say and talk about thingswhich we cannot understand (define).\qpa{I say more than I mean.I do not mean everything I say but I say everything I mean.}\pa{\em Self}-expression is an expression with bad conscience, an expression which knows and admits: \wo{this isn't really so, it is so only {\em for me}}.\paNot so seldom, wrong signs lead to right conclusions.\paNothing is ever definite except definitions (concepts, decisions).\paWhat can and what cannot be communicated doesn't depend on the subject matterbut only on the people involved. Given a wrong audience it may be equally impossiblefor a mathematician to explain a most clear proof as for an artist his impression or for a saint his vision. That higher things are harder to explain means only that itis harder to find people who can understand them.\pa That something can't be described. defined, said does not mean that it is not.\qpa{Symbol (conscious/reason) {\em is} the object -- only in a different context thanits perception or experience.}\paWhen I say ``soul'' I am not committed to admit the existence of sould. WhenI say ``soul'' I admit only that this is a reasonable way of speaking. Allour committements are epistemic -- the ontological ones are expressions ofeither limited perspective or intellectual arrogance.\paOften, and in the spiritual sphere always, the invisibly thin lines are the mostcrucial boundaries. \NOWY{Work}\plan{Work}\upa{The goals must remain hidden until we reach them.}\paOnly distance makes a relation possible -- there is no closeness without distance.\paWhatever one is, another can be as well. But one cannot be the other.No human being possesses a quality which another human being couldn't possess.\paWork brings you closer. So does suffering. They are necessaryexpressions of the distance. (Until it ceases to matter.)\paDistance is finite -- and absolute. (Work has no end.)\paThe distance from \LL\ to \HH\ is infinitely long. But once you havecrossed the border it disappears.\paYou will find yourself in everything you do. But only there?\qpa{Success is an indication of a lack of content -- at least a lack of depth.}\upa{Man gains depth realizing that some questions should not be answered-- even if they have to be asked.}\paConcrete is what is close, not what is precise.\paWork is what keeps heaven and earth together.\paWork is the (a) process through which spirit gains control over the ``world''.\paMost significant changes (affecting the \HH) are effected by long series of acts and works none of which has in itself any apparent effects.\paTo create is to convince - first oneself, then others.\paRationality is a movement upwards under the spell of visibility: with thepre-given intuition of unity, an attempt to re-construct it as a totality(without ever surpassing the boundary of the visible).\paWhat hides behind is what lies ahead -- we do not know exactly. Or, slightlydifferently: ``Know what is in front of your face, and what is hidden fromyou will be disclosed to you.'' [Gospel of Thomas, 5]\pa Some scholars are like men who insist the more on their ability to swim, thecloser they are to drowning.\NOWY{Evil and/or pain}\plan{Evil}\qpa{There is no evil without suffering.}\paThe only basis of spiritual power is the ability to suffer, that is humility.\pa\wo{Passivity increases our chances for evil, for becoming evil.} Is it a triviality? Yes -- but only in an evil world.\paSin is its own punishment - it immediately perverts one's being, diminishes humility, openness, and thus, the enjoyment of the world.This, however, does not mean that it is immediately \LL. It may taketime before the sin is punished in a visible way. Indeed, it may happen that it never will. \pa``Chciwo\'s\'c (avaritia) jest s{\l}u\.zeniem ba{\l}wanom. Avarice is serving the idols.\paOnly those looking for heaven get themselves occasionally into hell. The twoplaces are terrifyingly close to each other, separated only by a hair-breadth.(Piek{\l}o jest wybrukowane dobrymi checiami.)\NOWY{Feel, Love}\plan{Feel, Love}\paThe first form of love is the need to be loved. Disregarding it we sow theseeds of quiet suffering which becomes evil.\paTo feel is to be vulnerable.\paNot returning love is to kill it.\pa``I hate you''. Is it cruel? Not at all -- it can be changed. ``I donot love you'' does not hurt more than honesty should do. But what isterrifying, disclosing the cruel irreversibility of time is to hear``I loved you ... ooh how I loved you.''\paWhat is true is like love -- it lasts by enriching everything itencounters; it does not negate anything. It is humble.But it is not defensive, saying ``it is good for {\em me} but,perhaps, not for you''. Defensiveness always hides some untruth.\paMi{\l}o\'s\'c rozgrzesza.\paPerfect, all inclusive egoism: Whatever you do, you do for your own sake -- because you find yourself in everything and everything in yourself.\upa{One regrets the lost intensity of feelings which created the impression oflife. Shouldn't one then try to listen to the most weak and feeble `feelings'which, lasting but not overwhelming, are the matter of whole life?}\upa{There are things which enter our experience only as longing and it is a greatmistake to think that it is longing for something. A true, original longingis not longing for ... -- it is simply longing, without any object, whichdiscloses reality beyond objects.}\pa\label{qpa:lin}Love is more than a feeling. It is readiness to meet with humility whatevermight come. And be thankful for it. (cf.\refp{qpa:in}).\paLove: those people offering each other their uncertainities, and then risingstone and rock solid houses out of them.\NOWY{God, Spirit, religion}\plan{God, Spirit, religion}\paEternity is the \co{presence} of the \co{absolute} in the temporal. Immortality is \co{concreteness} of this \co{presence}, that is, when this\co{presence} not only surrounds but also permeats \co{actuality}; not when itis felt (it never is), but when it is known and \co{recognized}. It can be\co{experienced} but is never \co{an experience}. Perhaps, wecould say that the moment of death is the only possibility of \co{an experience}of immortality, the only moment when the \co{presence} of the \co{absolute}becomes fully \co{actual}. (But do we want to say that?)\upa{Man is an animal with ontological hunger. If you like -- a religious animal.}\paHunger of being is a hunger of state -- of something that is, notsomething that becomes. \qpa{And state is something which isn't merely actual -- it must extend beyond the horizon of actuality. State is the first image of something that lasts.}\paWhat do you want to find looking for God beyond coincidences? Isn't Hethe One whose ways are veiled and therefore look like coincidences?What more do you want than just to accept the coincidences ...?\paReligion does not address the paradoxical. It addresses the paradoxical {\em in life}.\paCan you say anything meaningful which is beyond you control? Yes, youcan. And beyond your spirit?The limit of your spirit is where your words cease to have meaning. \qpa{The limit of your intellect is where your words cease to havemeaning. The spirit exists where the dreams begin.}\paIt is the spirit that has me -- not I who have spirit.\upa{Our vocation is to listen -- not to talk.}\paSpiritual acts are directed onto \HH, even if their object is anactual thing. Communion consists in co-performing and co-understandingsuch acts. What distinguishes people from animals is not that animalsdo not have spirit (how could we know?), but that we can experiencefull communion with people only. With animals we can at best communicate.\paCan a life void of prayer be dedicated to God?\paQuestions ``What is God?'' or ``What is He like?'' make no sense. Theonly meaningful question about Him is ``How should I relate toHim?''. Then His attributes cease to be properties and turn out to bemere analogies implying appropriate existential attitude.\pa\label{ugood} (cf.~\re{happylot})Balance and harmony are impossible as long as one nourishes theconviction that the world is not ultimately good.Admitting its indifference may be an act of intellectual honesty but,equally, it is an act of existential despair.\paTo be is to participate in something which transcends me. Being is participation.\paEverything that is is {\em in-between}. Even solitude andunderstanding are just forms of communication.\paFaith looks for understanding. Well, what does not? (Life is to make \HH\ present, even visible.)\NOWY{Life, Person}\plan{Life, Person}\paWhen every casual gesture, charming laughter, every innocent sign carryingunknowingly a promise of thepossible happiness is taken as falshood, as a deceitful promise of animpossibility -- then one's life is broken. \paWise man thinks about future, but does not make plans.\upa{To hope is to have no expectations. Hope is the infinite patience of waiting for the \co{invisible} whic is, even though it never happens. \citf{For we are saved by hope: but hope that is seen is not hope:for what a man seeth, why doth he yet hope for?But if we hope for that we see not, then do we with patiencewait for it.}{Romans VIII, 24-25}}\upa{Man is a borderline between the \LL\ and the \HH.}\upa{His life is making the \HH\ present. It's meaning is to bring and keep heaven and earthtogether %(cf.~\refp{pa:heavenearth}).}\paYou will never have anything which you earlier did not sacrifice. (Not even yourself,)Never desire anything you cannot accept not having.\paMistyka normalno\'sci. Mistyka racjonalizmu to chorobaniemo\.zno\'sci; racjonalizm mistyki -- nie.\upa{This is the great adventure of meeting new people -- they can makeimpossible look obvious and natural.}\paLife unfolds through extremes -- but persists through mediocrity.\qpa{Cz{\l}owiek jest tym, co po sobie zostawia.}\paWe have not decided what we are. But we are responsible for it -- forwhat we have not created.\paWe are not masters of our being.Yet, we have the power of acceptance and denial, the power to striveor withdraw, of rejecting or staying faithful (to what we could be).\paEvery explanation must stop somewhere. It tells you a great deal about a person if you notice where his explanations stop.\paEvery doubt can be resolved if one forgets it.\paYou want something -- try to get it. And if you cannot - try to forgetit. If you cannot forget it, not immediately but after months andyears, then it becomes yours.The real is what we cannot live without. Also what we cannot stopbelieving in and longing for.\paTo want something is to immortalize it. [Like to hope; to wait; ...]\upa{The question which the ideal theory of personal (subject's) identitymust answer in negative is ``Is the disintegration of personalidentity, the loss of one's inner source of being, motivation,consistency, continuity possible?'' But this answer is obviously wrong!}\pa\label{happylot} (cf.~\re{ugood})It is the prerequisite of a happy lot -- to turn disasters and miseryinto advantageous experiences. (And it is something very different from``learning from experience.'') \pa\label{goodevil}God creates from nothing. Human creativity is to derive good from evil. [Tischner]\qpa{What is beautiful is impossible -- and what is possible isn't interesting.}\paNever forget -- we walk on hell, gazing at flowers.\paThe ultimate consequence is typically the same as the bitter end. \\Extreme boredom becomes a bit funny.\\(Problemem prznikliwo\'sci nie jest, \.ze nie idzie do ko\'nca, lecz \.ze go mija.)\paA man who has no expectations is unable to hate. Yet, he can be able to love.\paJeg er ikke useri{\o}s -- jeg bare ikke tar seri{\o}sitet s{\aa} alvorlig.\paIn principle, everything is possible but in practice, almost nothing happens.\pa The phrase ``commonly accepted rules (of conduct)'' means nothing more than predictability.\paPrinciples which some people claim to follow or, at least, to have, are completely different from the principles and laws of science: the latterare designed so that almost all phenomena conform to and agree with them -- theformer are so that almost none do.\upa{ Decision, being an act of reflexive consciousness, externalizesits object, that is fixes/freezes it.  Thus it may last, independentlyfrom our feeling s and state of mind, supported only by further actsof will.}\upa{What makes it difficult to accept (hard) truth is the faith, the hope that it is false. It isn't a lack of realism.}\paDo not defend it, do not fight for it, do not try to reach it -- just do it.\upa{Modesty depends on one's standards. If they are too high, others will rather see ambition and pride.}\paNothing helps more for self confidence than to ignore the possibility of making mistake.\pa\wo{Two weeks in Prague with my girlfriend}, although an individual event, was common -- nay, {\em shared} between the two of us. Life, when it is shared, is neither yours nor mine. I do not share my life with you, you do not share your life with me -- we share life, {\em the} life.\paArrogance is the usual end of false humility.\paA man is a dream -- his invisible dream. Don't say it -- dream it!\upa{Do not search -- find! If you start searching, you will hardly find.}\paIf you think that \co{separation} has, essentially, some negativeconnotation, then consider that it is the only event signifyingtranscendence, otherness, that is, the only event enabling relation,abolishing loneliness. If everything is mine -- or myself, if I am notseparated from anything, then I am alone.\NOWY{``Morality''}\plan{``Morality''}\paIt is impossible to reatin any self-value (self-respect?) maintainingnihilism towards the rest of the world. For nihilism is rejection ofcommittment, while it is only through committement that I can acquire -- thatI acquire -- a value, become worthy. To committ onself, to take upresponsibility is to say: ``I am worthy of carrying this responsibility.''Committement is the only way of (living) recognition of the (value of a)thing, person.\paResponsibility is the stronger, the more we feel that the thing, the person, whatever, for which we are feeling responsible is not us, is not ours. On the one extreme, it is a \wo{moralistic} attitude of freezing, impersonal responsibility. But, on the other extreme, it is just an aspect of accepting a thing as us, as ours, and with it, the responsibility for it which is the same as the responsibility for ourselves. Here, the word \wo{responsibility} is no longer adequate. There is no longer \thi{something to be responsible for}, only something which we take as seriously as ourselves, an aspect of ourselves deserving the same care, or lack thereof.\NOWY{Time}\plan{Time}\paFor the young days pass quickly but years slowly; for the old days pass slowlybut years quickly.When you have seen or experienced years of friendship, longlasting love which gradually, over the years, deteriorates and forgets itself,when you have seen people being born and die -- what possibly could then happenduring one day, one hour? The moments lose their weight as the significantexperiences stretch over months and years, when the \thi{essence} of last 20-30years gets condensed to a few observations, perhaps, just a single impression...\paTime is the means by which we can value things. [The only way to endow thingswith value is to devote them one's time.]\paI do not have enough time to be in a hurry.\paTime never goes back. It moves on and on, forever -- until you stop it.\paA moment devoid of awaiting for eternity becomes a moment of desperate awaiting forthe next moment. \\(And because a single moment does not bring anything, the hope fades away.)\pa``One can think that authentic time is originally an ecstasis; yet one buysoneself a watch.'' [Levinas]\paA man who never dwells on the past in order to always move forward, nevermoves an inch.\paYou can't remove your past -- the only thing you can do is to change it.\pa\co{Sign} is the result of a duration (continuity) exisiting through\co{actuality} (ontological perspective) -- every \co{sign} is a trace...\paIt is no big art to realize that one is a fool, but it is a true art to stop being one. We do not cure a disease by pronouncing the name of a medicine, but by taking it.\citf{No one can be made perfect in one day.}{Theologia Germanica XIII}\NOWY{Freedom, Others, Solitude}\plan{Freedom, Others, Solitude}\pa``Humility is a way of being oneself.''\paSilent emptiness is not when nobody is speaking but when nobody is listening.\pa``Bo ja tam ko\'ncze\; sie\;, gdzie mo\.zno\'s\'c moja'' [Norwid]\refp{goodevil}\paYou aren't free unless you can say ``the world is mine''.Not ``this is mine'' or ``that is mine'' but ``the world is mine''. Itis mine because it is the way it should be, because I accept it fromthe bottom of my heart. To have something means to accept it the wayit is from the bottom of one's heart.\paThe opposite of solitude is not sociality; it is personal contact withother human beings, friends, family. The oposition solitude-sociality ismisuderstanding coming from socilogical opposition private-public life.Society can never solve the problem of loneliness.\upa{``One day man will go mad to prove that he is free.'' [Dostojewski]}\upa{To be nothing is to owe everything (to others). Everybody, also ultimate freedomowes most it has. One remembers the unhappy events and days of childhood, complaining about familyand relatives who did not do their due... About the society which did not anddoes not function in the way promoting personal happinness... And about others, met thenand now, who take aways one's spare time, money, possibilties of enoyment, one'slife. All such complains my seem justified but their only work is: enslavement. Nobody owes anything to freedom.}\upa{\label{flirt2}Thirsting for the woods [origin, begining], we rise cities [construct, reason]. %}%\upa{\label{flirt}We flirt with time, yearning for eternity. %\refp{flirt2}%}%\upa{Thirsting for eternity, we flirt with time. %\ref{flirt}}\pa You are threatened only to the extent you have something toprotect.  And you are the more threatened, the more responsibility youassume for protecting what is yours.  [Exaggerated responsibility isan arrogance.]\NOWY{Understanding}\plan{Understanding}\paWe believe to understand other persons mostly by imputing them motiveswhich would be acceptable to us. \paBeing \thi{ahead of one's time} involves being out of one's time. Society prisessuch people (at least those of them, who happened to say something fitting the{\em current} situation) but one should rather look at their lives...(and what?)\paFundamental principles (of philosophy) never explain details!\paWhich distinctions one is able to make and consider tells something about one's intellect. Which distinctions he does not consider tells more about himself.The more we think, the greater the need for distinctions. We begin to be once we stop distinguishing.\paUnity is the horror of the intellect.\paOnly illiteracy allows one to nourish the hope that {\em the} truthhas been written down somewhere (else).\paThe fall begins with explanations, the belief that everybody should,or at least could, understand.Except at the universities, all popular explanatory writings should be forbidden.\upa{Arguments never convince.Argumentation is a sign of lacking respect - one tries to convince by explaining toanother what he apparently is unable to understand. If he only could, he wouldaccept our conclusion.}\paNever accept an attitude or a view which you find implausible only because you cannot find a counterargument for not accepting it.\upa{There is nothing unreal. How could there be? It takes a lot ofdisappointment to rise suspicion, and a lot of suspicion to claim thatthe reality consists of two parts: ``real'' and ``unreal''.}\upa{``Reality'' was invented as a medicine against insecurity anduncertainty, as a reaction against one's failure. And immediately,there followed the search for its infallible criteria.}Thus epistemology was born out of suspicions against ontology whichwere nothing else than suspicions of one's own failures.\pa\label{anymisunderstood}There is nothing that could not be misunderstood. It is just anotherside of the fact that it is so difficult to say something which makesabsolutely no sense.\pa It is better to be not understood than misunderstood.  \paWisdom is the art of hiding -- and simplifying.\paWisdom does not make saints.\paTo be wise is not to know what one does not know but to know whatcannot be know.\paCreate order: simplify the complicated and complicate the simple.\pa``Life is simple but man insists on making it complicated.'' [Confucius]\pa``He that hath ears to hear, let him hear''.[Mat.11:15]\paIdealism is an unrealized illusion -- or an unfulfilled reality.\paExplanation is a reduction. This is what makes it so interesting and, from the existential point of view, so irrelevant.\paIn order to say something understandable we have to create a world.\qpa{To speak is to create.}\paThe role of distinctions isn't to separate but to unify.\paAmong all the things which can be distinguished almost none can be defined.\paWhere is the borderline separating facts from feelings? But does the fact that it cannot be sharply defined mean that it does not exist?\paExperience is the best reason to change a view.\paDo not try to explain to yourself -- and especially to others -- what and why you aredoing. For the most, you do not know. And almost inevitable misunderstanding will lead youinspiration astray. You'll understand when you have done it -- when you have become mature.\upa{To explain is to reduce. To understand is something very different.}\paThe problem of spiritual understanding is not that it is hard to understandbut that it is so easy to forget what one understood.\pa Truth is agreement with the origin. (At each level, it may be the local origin of the sign for the local correlate.)\paIt is one of the charms -- that contradictions which reflection so vehemntlyforbids to be and not to be at the same time and place, co-\co{exist}, oftenquite peacefully, sometimes a bit violently...\NOWY{Certainty}\plan{Certainty}\paOnly the fear -- the fear of the unexpected -- makes us obsessed with certainty.\pa It is a common fallcay: the possibility of non-$A$ proves non-necessity of $A$ andis immediately taken for $A$'s non-reality.\paYou want certainty? -- you have to lose even before you start playing.\\You want to win? -- you have to forget that already at the start you lost.When the heart weeps for what it has lost, the spirit laughs for what it hasfound. [anonymous Sufi aphorism, after Huxley's Perennial Philosophy, p.106]\paDefeat is the only source of strength. For to be strong is to be unafraid ofdefeat (even of death). You are strong when you have learnt that it isimpossible to lose, no matter what defeat you suffer. \pa Anything that can be achieved by means of (operational)understanding and (rational) control is suspicious -- that is,uncertain.  For if one can arrange circumstances so that it appears,then somebody else may arrange them differently so that it disappears. The more things are liable to manipulation, the more relative, changeable and, eventually, uncertain they become.The only certain thing is the one which cannot be obtained by manipulation: grace --the undeserved, incomprehensible gift.\paCommitments come first -- responsibility is only a consequence.\qpa{Kiedy pojawiaja\;\ sie\;\ pytania, na kt\'ore nie ma odpowiedzi, oznacza to, \.ze nasta{\;}pi{\l} kryzys. [R.~Kapu\'sci\'nski]}\paNever look for contradictions -- they do not exist.\paThe most subtle art of contradictions.\NOWY{Meaning}\paTo live is to perceive things as meaningful.\paAll or nothing: either everything has a meaning and significance or nothing does -- eventually.\paIt isn't true that when you have a goal, perhaps even many goals, your lifeacquires meaning. Often the doubt and questions about the meaning appear exactlyin the middle of hectic activity stimulated by several goals. Goals can, at best,make you not feel meaninglessness.\paSpirit is the meaning of things. I do not create it -- it is given to me ... or not.\pa``Everything is possible'' means ``the very best, highest can happen''.\paEven things are expressions; indirect and imperfect but neverthelesssignificant -- perhaps, the only expressions we have. This is what makes them meaningful.\paMovement and meaning of the world concentrates in things -- dying out. (zamieraj{a}c)\NOWY{Levels}\pa\HH\ in \LL: metaphor, contradiction, unclarity; \LL\ on \HH: confusion, misapprehension, contradiction.\paThe higher the spiritual level, the fewer possible things. Not because they are impossible but because they became unacceptable.\pa\label{qpa:in}Higher (or deeper) than \co{things} and concepts are feelings. But stillhigher are things which we hardly can call feelings, things which we merely-- and simply -- know (\co{invisibles}, cf. \refp{qpa:lin})\upa{The relation between higher and lower, invisible and visible is not that ofthe general and the particular, that of instantiation and specialisation. Itis the relation of expression, of incarnation. Framing it in terms of generality and specificity, the tradition hascompletely disregarded its nature, casting it in merely conceptual terms.}\upa{It was the assumption that the higher, eventually the highest, somewhat`contains' everything lower, that forced one to double things with ideas orelse, to make God responsible for all the details of this world.}\NOWY{Truth}\paOften, most false assumptions may lead to most true conclusions, just like usually the true assumptions will yield most unbearable conclusions if only arrived at with sufficient degree of correct analysis and unbearable patience of reason \ldots \paThere is a major difference between the possibility of \thi{me being mistaken}and \thi{the thing being something in itself}. I can be mistaken because thething can be seen from a different perspective, considered in a differentcontext, used for another purpose. This, in turn, is very different from\thi{thing being only the perspectives on it}.The whole problem is with the transition between the \co{actual}, \co{visible}and the \co{invisible}; in the more mundane words, between what is consciouslychosen and what is given for consciousness.Thing is only a limit of distinctions, it is indeed relative to thedistinguishing being, an existence, but this \co{existence} is more than ourbeloved \co{I}, not to mention \co{subject}. Thing is constitutedtranscendentally, only that this transcendentalism is of a pretty concrete, notto say empirical, kind. It involves my/human brain and perception mechanisms asmuch as, or rather more than, my voluntary choices. \thi{Perspectives} are notchosen -- they are given. They are given as much as the thing itself is given to consciousness. \co{I} cannot choose another perspective on it than those which are given with it, whichare inscribed into the distinctions constituting the thing. Indeed, things aregiven us to the extend they are not underlied our control, to the extend theykeep something away. And as we dissolve more and more things, as we dissolvemore and more the apparently solid world around us, we ask more and moreintensely what really is there, where is this whole reality we have though sosolidly given and determined?\paIf we were to define it, we would say that truth is (agreement) with\co{origin}. (At each level, \co{origin} can be understood differently.)\paWe do encounter examples of genuinely emotional reactions as well as ofhighly detached mental attitudes. This, however, does not mean that there isany line separating feeling from thinking, that they do not relate to thesame reality.\paTruth is confusing. I mean, search for truth and only truth breedsconfusion. For even if not everything is truth, everything is at least anaspect of truth, and one is soon drawning in the proliferation of trueobservations, true remarks, true statements. The question is not ``What istruth?'', not even ``What is true?'' but ``Which truth?'' (shall I live).\paIf you want to feel thinking, or think feeling, Being, try to think something which can be dissociated but which is not, some aspects which, although in principle separable, are in fact inseparable -- not because of any necessity, not because they {\em must} come together, but simply because they do.Insight of identity of two things which we have considered distinct is but a meagre analogy of that \ldots\NOWY{Women/Men}\paMen attempt not to fail. (The least success is the lack of failure.) Womentry not to miss the truth. Or else: men are satisfied when they don't fail,women -- when there is nothing missing.\paBeing concretely present means being close to the chaos; not overwhelmed anddefeated by it, but close to it, constantly threatened by it. For chaos --the uncontrollable flux of reality -- is the ultimate, concrete content oflife and of each situation, suspending the separation of the external fromthe internal. Therefore women, yielding to the insecurity of the fluctuationsof their moods and feelings, are much more concretely present than men are.This, however, is only statistical and general truth, so to speak,concerning the given nature.  In exceptional cases, it is typically aman who can be \co{present} with the irresistible force of ultimate\co{concreteness}.  Such \co{presence} is \co{founded} beyond\co{chaos} of mere feelings, it is a rare \co{spiritual} event.\NOWY{Letter of Count (hrabia) Joseph de Maistre}to Natalia Wiaziemcow (fiancee of his son, Rudolph de Maistre), Spring 1813.[Tygodnik Powszechny, No.43, 27.10.2002,  ``List Hrabiego'' w: Apokryf na75-lecie Leszka Kolakowskiego] \paMasy ludowe szanuja rzad tylko dlatego, ze nie jest ich dzielem. Ulegajazwierzchnosci, bo czuja, ze to cos swietego, czego nie moga ani stworzyc, anizniszczyc.\pa... jest przeznaczeniem zwierzchnosci karanie zla i z tego powodu wladza jestnieomylna w kazdej dziedzinie, w ktorej zlo sie ukazuje, a ukazuje sie wkazdej.\paCala wielkosc, cala potega, cala dyscyplina spoczywa na kacie: on jestpostrachem i wiezia spolecznosci ludzkiej. Zabierzcie swaitu ow niepojetyelement, a w jednej chwili lad ustapi miejsca chaosowi, trony upadna ispoleczenitwo zniknie.\paZdolnosc do samoograniczenia sie jest dowodem pojmowania dystansu. A to zpoczucia dystansu, z {\em ironii} wlasnie, ktora tak pogardzasz, rodzi sieusmiech [\'{s}niony mlodzieniec do hrabiego]%\newpage\NOWY{St.~Augustin}\citt{Ojciec karci syna, handlarz schlebia niewolnikowi. Gdybys obydwa czyny --kare i delikatnosc -- dal do wyboru, ktoz by nie wybral delikatnosci i nieodrzucil bicia. Niejeden czyn wyglada na zewnetrznie dobry, choc nie wyrasta zkorzenia milosci. Podobnie i kwiaty maja ciernie. Niejeden czyn wyglada nasurowy i twardy, choc dokonuje sie go dla wychowania, z pobudki milosci. Dlategopolecam ci jedno krotkie zdanie: Kochaj i czyn, co chcesz!}{Homilia na 1.Listsw.Jana (hom. 7,8) [Love and do what you want!]}\la{Qui bene amat, bene castigat} (Kto kocha wlasciwie, ten nalezycie karci.)\NOWY{Dostojewski, Biesy}\paByla to mysl cyniczna, lecz ludzie o duszach wznioslych czasem nawet majatendencje do mysli cynicznych, chociazby tylko dzieki wszechstronnosci  swegorozwoju. [I.1.iv] [Stiepan Wierchowienski, p.22]\paWolnosc nastapi wtedy dopiero, gdy juz bedzie wszystko jedno: zyc czyumierac. Oto jedyny cel. [I.3.viii] [Kirillow, p.119]\paOd prawdziwego cierpienia madrzeli nieraz glupsi, nie na dlugo,oczywiscie. Cierpienie ma juz taka wlasciwosc. [I.5.vii] [ad StiepanWierch. p.303]\paSa mysli wieczne, ktore nagle staja sie nowymi. [II.1.v] ] [Kirillow, p.237]\paTak juz wszystko jest nudne, ze nie ma powodu przebierac w rozrywkach, byleby tobylo cos ciekawego. [II.5.ii] [Jedna z panius, ktora chce tez zobaczyc cialosamobojcy, p.325]\upa{Nihilistami w istocie rzeczy bylismy my, wieczni poszukiwacze ideinadrzednej. Teraz rozmnozyly sie badz zobojetniale miernoty, badzmnichowie. [Dostojewski w notatnikach, p.725 [18]]}\pa\fre{Libert\'{e}, \'{e}galit\'{e}, fraternit\'{e}}\paJezeli nie ma Boga, to ja jestem bogiem. [III.6.ii] [Kirillow, p.610]Uswiadomic sobie, ze nie ma Boga, i nie uswiadomic sobie jednoczesnie, ze siesamemu jest bogiem -- to nonsens. [III.6.ii] [Kirillow, p.612]Jesli jest Bog, to i ja jestem niesmiertelny. [III.7.iii] [chory Stiepan, p.656]\paWiedzialem o tym od dawna, lecz widze dopiero teraz. [chory Stiepan, p.645]\upa{Tutaj najmniej lubilem mieszkac. Ale nawet tutaj niczego nie moglemnienawidzic. [...] Moge zechciec zrobic dobry uczynek i robi mi toprzyjemnosc. Ale juz zaraz pragne zlego i tak samo czuje przyjemnosc. I to, itamto uczucie jest jak zawsze zbyt plytkie, a nigdy nie chce bardzo. [III.8.epilog] [Mikolaj Stawrogin, p.666-7; ad. Sade]}\paWszelkie sytuacje niezwykle hanbiace, nadmiernie ponizajace i, co najwazniejsze,osmieszajace, w jakich sie w moim zyciu znalazlem, zawsze wzbudzaly we mnie narowni z bezmiernym gniewem, niewiarygodna wprost rozkosz. [U Tichona][Stawrogin, p.682 - cf. Rousseau [9]; ad. Sade, silne przezycia]\upa{Wiem, ze powinienem zabic siebie, zmiesc siebie z powierzchni ziemi jakszkodliwego owada. Lecz boje sie samobojstwa, gdyz boje sie okazacwspanialomyslnosc. [III.8.epilog: Stawrogin, p.667; ad Th.Germ. self-despise]}\paPragne, aby bylo wiadome, ze ani we wplywie srodowiska, ani w chorobie niezamierzam szukac usprawiedliwienia dla moich zbrodni. [U Tichona] [p.683]\upa{Kazdy czlowiek grzeszac, grzeszy przeciwko wszystkim ludziom i kazdy czlowiek wjakims stopniu winny jest cudzego grzechu. [U Tichona] [Tichon, p.699] }\paTu wydarzylo sie to, o czym pisze Ewangelista Lukasz [.. Lk.VIII:32-36..]Biesy wyszly z czlowieka rosyjskiego i weszly w stado swin, to znaczyNieczajewow i innych. Potonely one lub potona na pewno, a uzdrowiony czlowiek, zktorego wyszly biesy, siedzi u nog Jezusowych. Tak i byc powinno. Rosjazwymiotowala te ohyde, ktora ja karmiono, a w owych zwymiotowanych nikczemnikachnie ma juz, oczywiscie, nic rosyjskiego. I niech pan zapamieta sobie,przyjacielu: kto traci swoj narod i narodowosc, ten traci i wiare ojczysta, iBoga. Jesli pragnie pan wiedziec -- wlasnie to jest tematem mojejpowiesci. Nazywa sie one Biesy is stanowi opis tego, jak biesy weszly w stadoswin. [Dostojewski: list to Apo{\l}{\l}ona Makowa z Drezna, 9/21.10.1870]Lk.VIII: [32] And there was there an herd of many swine feeding on the mountain: and they besought him that he would suffer them to enter into them. And he suffered them.[33] Then went the devils out of the man, and entered into the swine: and the herd ran violently down a steep place into the lake, and were choked.[34] When they that fed them saw what was done, they fled, and went and told it in the city and in the country.[35] Then they went out to see what was done; and came to Jesus, and found the man, out of whom the devils were departed, sitting at the feet of Jesus, clothed, and in his right mind: and they were afraid.[36] They also which saw it told them by what means he that was possessed of the devils was healed.\NOWY{Herbert, \btit{Kr\'{o}l mr\'{o}wek}}\pa\wo{chtoniczny} = \gre{chtoni\'{o}s}, zrodzony z ziemi, podziemny\paLos i lokaj losu -- przypadek. (Fate and its lackey -- chance.)\ad{Narcyz}...poszukuja daremnie sensu \.{z}ycia na dnie pustej duszy... [p.63]Jednakze brutalno\'{s}\'{c} i g{\l}upota potrzebuja spoiwa, trzeciegopierwiastka, by stworzy\'{c} trwala moleku{\l}e charakteru. Tym pierwiastkiemjest najcze\'{s}ciej sentymentalizm. Wiec i on zakocha{\l} sie. [p.63]Uroki opieku\'{n}stwa zblad{\l}y. [p.64]\ad{Kleomedes}Ale  katastrofy sa przecie\.{z} niewinne. [p.55]Poni\.{z}enie daje pewno\'{s}\'{c}. [p.57]\upa{(ad. Contradictions)\begin{quote}Odi et amo. Quare id faciam, fortasse requiris. \\Nescio, sed fieri sentio et excrucior.\\(Kocham i nienawidze. Jakze sie to dzieje? Spytasz.\\Nie wiem, lecz czuje, cierpie i szaleje.)\end{quote}[Katullus, \btit{Poezje, Piesni:85}]}