\ad{distinction but no essence}We can imagine a \co{distinction} without \co{ipseity} when we encounter something for the very first time: we do not re-cognize it, we do not know, nor even feel, what this thing, this feeling, this something might be. But we know that it is {\em something}, only that we do not know what, we encounter it as something different from everything else.\ad{holiness, concepts}Before you meet a holy man, you may have a very confused, if any, concept of holiness. But it is no goal to have all the concepts correct and ready in advance. The goal is only this: {\em when} you meet a holy man, do not let your concepts (or misconceptions) prevent you from recognizing his holiness, from admitting it. (Actually, concepts, no matter what the hermeneutics tell, make it harder to accept the testimony of new experiences.)\ad{holiness, the Other}Holiness is never experienced as the Other, the {\em absolutely Other} (unless one denies it). On the contrary, it embraces everything around itself, gives an unmistakeable feeling that \thi{also I am a part of it}.\ad{actuality confused with eternity}One of the main effects common to all the drugs is that they reduce the world to the pure \herenow, they narrow the horison of perception and consideration to this very moment, without tomorrow, without yonder, to pure actuality so seductively resembling the mystical presence of eternity.\ad{Analytical philosophy}says \wo{this is bubbling, incomprehensible rubbish}. Well, if I read only one page from Kant, it is incomprehensible, but it helps if I read more and think about it. When I read Bible for the first time I understood much less than when I read it again. But so was it when I read Carnap, and when I studied mathematics!\ad{expression vs. expressed}The question about truth has been all the time confused with the question about the formulation of truth; the absolute with the expression of the absolute. And finding no unique, unequivocal, verifiable formulation, one decided to abandon the quest.There is a deep lack of respect, both for the truth and for the other in this. The truth need not -- {\em can not} -- be expressed in a single way like a mathematical definition. (It is abundant, overflowing, non-actual.) And the other can understand the intention beyond the partial, even misleading expressions.\ad{reflection}The art of reflection is not to jump to a higher level of self-consciousness and observe everything, oneself included, from above and outside. On the contrary! It is to stay inside and from within to record the events, feelings passing by without externalizing them. It is to realize that there is nothing outside, that everything is and must be seen only from within. The art of reflection is not to oppose immediacy but to preserve it -- in spite of reflection.\ad{Community-personality}\ad{faith-ateism...}The alternative, the contradiction \wo{God is} vs. \wo{There is no God} ispossible only after one has reduced the supposed \thi{being of God} to thelevel of \co{actual experience}, to the level of \thi{being a thing}. It helpsnothing to claim that this was not the intention, for this is, in fact, theresult; and when the results are clear, the intentions do not matter. Assumingthat such an alternative is at all possible, one has already falsified themeaning of the most well intended and positive answer to the question \wo{Is there God?}.A better question would be \wo{What does \wo{God} mean?}, or even \wo{What does  God mean?}, though this, obviously, involves one in the matters which even themost prominent theologians eventually had to give up. Any answer could be accusedof arbitrariness -- but this should not worryThe advantage is that the previous alternative (\wo{He is} vs. \wo{He is not})now becomes \wo{He means nothing} vs. \wo{He means something}. The first onewill quickly declare Him to be non-existent. But this is now fine because nowit is also more clear that it is only inability, perhaps even only \co{my}inability, to find any meaning in Him (in it?), which leads \co{me} to thisconclusion.If there is any God, He can be only in my life, and if He is not there then,indeed, He is not...