\documentstyle{report}

\input{apemacro}

\def\apekeywordfont{\footnotesize\bf}
\def\apecommentfont{\footnotesize\it}
\def\apebodyfont{\small\rm}
\def\apelmargin{0.5cm}

\setlength{\textwidth}{6.8in}
\setlength{\textheight}{8.7in}
\setlength{\topmargin}{-1in}
\setlength{\headheight}{0in}
\setlength{\headsep}{1in}
\setlength{\columnsep}{.3in}
\setlength{\oddsidemargin}{-0.15in}
\setlength{\parindent}{0pt}
\setlength{\parskip}{\medskipamount}

\begin{document}
\pagenumbering{roman}

\newdimen\indz
\newdimen\inda
\newdimen\indb
\newdimen\indc
\newdimen\indd
\newdimen\inde
\indz=5in
\inda=\apehspace
\multiply\inda by -3
\indb=\inda
\indc=\inda
\indd=\inda
\inde=\inda
\advance\inda by \indz
\advance\indb by \inda
\advance\indc by \indb
\advance\indd by \indc
\advance\inde by \indd

% Cover page
\setcounter{page}{0}

\newdimen\coverindent
\coverindent=0.8cm

\def\tfb{\Large\bf}
\def\tf{\Large}

\vspace*{2.5cm}

{\tfb
\hspace*{\coverindent}\strut THE STANFORD ADA STYLE CHECKER:\\
\hspace*{\coverindent}\strut AN APPLICATION OF THE ANNA TOOLS AND\\
\hspace*{\coverindent}\strut METHODOLOGY}

\vspace{2.8cm}

{\tfb
\hspace*{\coverindent}\strut Micha{\l} Walicki\\
\hspace*{\coverindent}\strut Jens Ulrik Skakkeb{\ae}k\\
\hspace*{\coverindent}\strut Sriram Sankar}

\vspace{2.8cm}

\hspace*{\coverindent}{\tf Technical Report:} {\tfb CSL--TR--91--488}

\hspace*{\coverindent}{\tf (Program Analysis and Verification Group
                            Report No. 55)}

\vspace{2.8cm}

\hspace*{\coverindent}{\tfb August, 1991}

\vspace{2.8cm}

\hspace*{\coverindent}\parbox[t]{5.9in}
{Micha{\l} Walicki was on leave from the University of Bergen, Norway,
and was supported by the Royal Norwegian Council for Scientific and
Industrial Research.
Jens Ulrik Skakkeb{\ae}k was on leave from the Technical University of
Denmark, Denmark.
Sriram Sankar was supported by the Defense Advanced Research Projects
Agency contracts \mbox{N00039-84-C-0211} and \mbox{N00039-91-C-0162}.
}

\newpage

% Abstract page
\setcounter{page}{0}

\vspace*{1.6cm}

\begin{center}
{\tfb
\strut THE STANFORD ADA STYLE CHECKER:\\
\strut AN APPLICATION OF THE ANNA TOOLS AND\\
\strut METHODOLOGY}

\vspace{1.2cm}

{\large\bf
\strut Micha{\l} Walicki\\
\strut Jens Ulrik Skakkeb{\ae}k\\
\strut Sriram Sankar}

\vspace{1.2cm}

{\large Technical Report: {\bf CSL--TR--91--488}}

{\large (Program Analysis and Verification Group Report No. 55)}

\vspace{0.7cm}

{\large August, 1991}

\vspace{0.7cm}

{\large
\strut Computer Systems Laboratory\\
\strut Departments of Electrical Engineering and Computer Science\\
\strut Stanford University\\
\strut Stanford, California - 94305}

\vspace{2cm}

{\large\bf Abstract}
\end{center}

This report describes the {\em Ada style checker}, which was designed
and constructed in Winter and Spring 1989--90. The style checker is
based on the Stanford Anna Tools and has been annotated using Anna.
The style checker examines Ada programs for ``correct style'' which is
defined in a {\em style specification language (SSL)}.
A {\it style checker generator} is used to automatically generate a
style checker based on a set of style specifications.

\vspace{10pt}

\noindent
{\bf Keywords}---{\it
Ada,
Anna,
style specification,
style checking.}

\newpage

% Copywrite page
\setcounter{page}{0}

\vspace*{2.8in}

\begin{center}
\Large Copyright \copyright\ 1991
\end{center}

\begin{center}
\large by
\end{center}

\begin{center}
\Large
\strut Micha{\l} Walicki\\
\strut Jens Ulrik Skakkeb{\ae}k\\
\strut Sriram Sankar
\end{center}

\newpage
\setcounter{page}{3}

\newcommand{\version}{1.20}

\tableofcontents

\chapter{Introduction}
\label{intro}
\pagenumbering{arabic}
\input{tech-introduction}

\chapter{The Style Specification Language}
\label{ssl}
\input{tech-ssl}

\chapter{System Overview}
\label{init-system}
\input{tech-init-system}

\chapter{Using The Style Checker}
\label{usage}
\input{tech-usage}

\chapter{The Programmatic Interface}
\label{prog-interf}
\input{tech-interf}

\chapter{Future Work}
\label{future}
\input{tech-future}

\bibliography{style}
\bibliographystyle{alpha}

\appendix

\newenvironment{stylerule}[2]
{

\vspace{6pt}

\Large {\bf Rule No. #1 - #2} \normalsize
}{}

\newcommand{\guideline}[1]{\large {\bf Guideline:} \normalsize \\ #1

\vspace{10pt}

}

\newcommand{\ssl}[1]{\large {\bf SSL form:} \normalsize \\ {\tt #1} 

\vspace{10pt}

}

\newcommand{\internal}[1]{\large {\bf Internal Representation:} \normalsize
\\ {\tt #1}

\vspace{10pt}

}

\newcommand{\descr}[1]{\large {\bf Description:} \normalsize \\ #1

\vspace{10pt}

}

\newcommand{\remarks}[1]{\large {\bf Remarks:} \normalsize \\ #1

\vspace{10pt}

}

\chapter{The Stanford Style Checker}
\label{stanford}
\input{tech-stan}

\chapter{Maintenance Guide}
\label{maint}
\input{tech-maint}

\end{document}
