\documentstyle[a4wide]{article}
%%\documentstyle[11pt,a4wide]{article}
% these pages are quite small. These are Springer LNCS pages for 12pt
%\textwidth 16cm \textheight 23.5cm
%%\textwidth 16cm \textheight 24cm
\voffset=-.5in
%\hoffset=-.7in

\input{defs}

\title{Complete Reasoning and Rewriting with Set-Relations}

%changes:  interlines spaces decreased
\author{ {\it Valentinas Kriau\v ciukas\thanks{Both authors gratefully 
acknowledge the financial support received from the Norwegian Research Council.}}\\[-.5ex]
\small Department of Mathematical Logic\\[-.5ex]
\small Institute of Mathematics and Informatics\\[-.5ex]
\small Vilnius, LITHUANIA\\[-.5ex]
\footnotesize valentinas.kriauciukas@mlats.mii.lt 
\and 
{\it Micha{\l} Walicki\/\(^\ast\)}\\[-.5ex]
\small Department of Computer Science\\[-.5ex]
\small University of Bergen\\[-.5ex]
\small Bergen, NORWAY\\[-.5ex]
\footnotesize michal@ii.uib.no}


\begin{document}

\maketitle
\abstract {The paper investigates reasoning with set-relations: intersection, 
inclusion and identity of 1-element sets. A language is introduced which, 
interpreted in a multialgebraic semantics, allows one to specify such 
relations. 
Each of these set-relations satisfies only two among the 
three properties of the equivalence relations - we study rewriting with such 
non-equivalence relations and point out differences from the equational case. 
The introduced complete inference systems
generalize ordered paramodulation and superposition calculi.  Notions
of rewriting proof and confluent rule system are defined for such
non-equivalence relations.
 Together with the notions of forcing and redundancy
they are applied  in the completeness proofs.  

In the first part, an inference system is given and shown sound and 
refutationally ground-complete for 
a particular proof strategy which selects only maximal literals from the premise 
clauses. 
As a corollary of the ground-completeness theorem we obtain 
ground-completeness of the introduced rewriting technique.

The second part studies rewriting with the non-ground clauses.
Ground completeness cannot be lifted to the non-ground case
because substitution for variables is restricted to
deterministic terms.
To overcome the problems of restricted
substitutivity and hidden (in relations) existential quantification,
unification is defined as a three step process: substitution of determistic 
terms, introduction of bindings and ``on-line'' skolemisation.
The inference rules based on this unification derive non-ground clauses
even from the ground ones, thus making an application of a standard
lifting lemma impossible. The completness theorem is proved directly without 
use of such a lemma.}


\noindent{\bf Keywords:}
rewriting, theorem proving, binary relations, nondeterminism, 
completeness, ordered superposition



\input{cslA}

\input{cslProof}

\input{adt}


\begin{thebibliography}{MM99}\MyLPar
\bibitem[Bez90]{Bez} M.~Bezem. 
   Completeness of Resolution Revisited. 
   {\em Theoretical Computer Science}, 74, pp.27-237, (1990).
\bibitem[BG91]{BG} L.~Bachmair, H.~Ganzinger. 
   {\em Rewrite-Based Equational Theorem Proving with 
                           Selection and Simplification.}
   Technical Report MPI-I-91-208, Max-Planck-Institut f. Informatik, 
   Saarbr\"ucken, (1991).
\bibitem[BG93]{BG249} L.~Bachmair, H.~Ganzinger. 
   {\em Rewrite Techniques for Transitive Relations.}
   Technical Report MPI-I-93-249, Max-Planck-Institut f. Informatik, 
   Saarbr\"ucken, (1993). [to appear in LICS'94]

\bibitem[BG95]{BG-Oslo} L.~Bachmair, H.~Ganzinger. 
   {\it Ordered Chaining Calculi for First-Order Theories of Binary Relations}.
   Technical Report MPI-I-95-2-009, Max-Planck-Institut f. Informatik, 
   Saarbr\"ucken, (1995).
\bibitem[BGLS95]{Basic-par} L.~Bachmair, H.~Ganzinger, C.~Lynch,  W.~Snyder. 
   Basic paramodulation. {\it Information and Computation}, 121(2),
   pp.~172--192 (1995). 
\bibitem[BGS94]{rel-spec} R.~Berghammer, T.~Gritzner, G.~Schmidt. Prototyping
   relational specifications using higher-order objects. In {\it
   Proc.\ Int.\ Workshop on Higher Order Algebra, Logic and Term Rewriting
   (HOA'93)}, Amsterdam, The Netherlands, Sept.\ 1993, LNCS, vol.~816,
   pp.~56--75 (1994).
\bibitem[BK95]{BK} M.~Bia{\l}asik, B.~Konikowska, {\it Reasoning with 
   Nondeterministic Specifications}, ICS PAS Report no.~793, Institute
  of Computer Science, Polish Academy of Sciences, (1995).

\bibitem[DJ90]{Der} N.~Dershowitz, J.-P.~Jouannaud. 
   Rewrite systems. In: J.~van Leeuwen (ed.) 
   {\em Handbook of theoretical computer science}, vol. B,
   chap. 6, pp.243-320. Amsterdam: Elsevier, (1990).
\bibitem[DM79]{DM} N.~Dershowitz, Z.~Manna. 
   Proving termination with multiset orderings. 
   {\em Communications of the ACM}, 22:8,pp.465-476, (1979).
\bibitem[DO92]{DO} A.~Dovier,E.~Omodeo,E.~Pontelli,G.-F.~Rossi. 
   Embedding finite sets in a logic programming language. 
   {\em LNAI}, 660, pp.150-167, Springer Verlag, (1993).
\bibitem[Hes88]{PS1} W.H.~Hesselink. A Mathematical Approach to Nondeterminism
   in Data Types. {\em ACM Transactions on Programming Languages and Systems},
   10, pp.87-117, (1988).
\bibitem[Hus92]{Hus} H.~Hussmann. Nondeterministic algebraic
   specifications and nonconfluent term rewriting. {\em Journal of Logic
   Programming}, 12, pp.237-235, (1992).
\bibitem[Hus93]{HusB} H.~Hussmann. 
   {\em Nondeterminism in Algebraic Specifications and Algebraic Programs.}
   Birkh\"auser Boston, (1993).
\bibitem[Jay92]{Jay} B.~Jayaraman. Implementation of Subset-Equational 
   Programs. {\em Journal of Logic Programming}, 12:4, pp.299-324, (1992).
\bibitem[Kap88]{Kap} S.~Kaplan. Rewriting with a Nondeterministic Choice
   Operator. {\it Theoretical Computer Science}, 56:1, pp.37-57, (1988).
\bibitem[KW94]{KW} V.~Kriau\v ciukas, M.~Walicki
   {\em Reasoning and Rewriting with Set-Relations I: Ground-completeness.}
   Proceedings of CSL'94, LNCS 933, (1995).
\bibitem[KW95]{KW} V.~Kriau\v ciukas, M.~Walicki
   {\em Reasoning and Rewriting with Set-Relations II: Completeness of the non-ground case.}
   in Recent Trends in Data Type Specification, LNCS 1130, (1995).
\bibitem[KW95a]{KMW}V.~Kriau\v ciukas, M.~Walicki.
   Nondeterministic Algebraic Specifications in Relational Syntax, in
   {\it Proc. of 7th Nordic Workshop on Programming Theory}, Report
   No.~86, Programming Methodology Group, G\"oteborg University, (1995).
\bibitem[LA93]{LA} J.~Levy, J.~Agust\'i. Bi-rewriting, a term rewriting
   technique for monotonic order relations. In {\em RTA'93, LNCS}, 
   690, pp.17-31. Springer-Verlag, (1993).
\bibitem[Mos89]{uni-al} P.D.~Mosses. Unified algebras and institutions. In
   {\it LICS'89, Proc. 4th Ann. Symp. on Logic in Computer Science},
   pp.~304--312, IEEE, (1989).
\bibitem[SL91]{relaxed-par} W.~Snyder, C.~Lynch. Goal directed strategies for
   paramodulation. In {\it Proc. 4th Int. Conf, on Rewriting Techniques and
   Applications}, LNCS, vol.~488, pp.~150--161,
   Berlin, (1991).
\bibitem[PP91]{PP} J.~Pais, G.E.~Peterson. Using Forcing to Prove Completeness
   of Resolution and Paramodulation. {\em Journal of Symbolic Computation}, 
   11:(1/2), pp.3-19, (1991).
\bibitem [S-A92]{S-A} R.~Socher-Ambrosius. 
   {\em Completeness of Resolution and Superposition Calculi.}
   Technical Report
   MPI-I-92-224, Max-Planck-Institut f. Informatik, Saarbr\"ucken, (1992).
\bibitem [SD86]{SD} J.~Schwartz,R.~Dewar,E.~Schonberg,E.~Dubinsky. 
   {\em Programming with sets, an introduction to SETL. }
   Springer Verlag, New York, (1986).
\bibitem[Sto93]{Sto} F.~Stolzenburg. 
   {\em An Algorithm for General Set Unification.}
   Workshop on Logic Programming with Sets, ICLP'93, (1993).
\bibitem[Wal93]{Mich} M.~Walicki. 
   {\em Algebraic Specifications of Nondeterminism.}
   Ph.D. thesis, Institute of Informatics, University of Bergen, (1993).
\bibitem[WM95a]{MW-II} M.~Walicki, S.~Meldal. Multialgebras, Power algebras
   and Complete Calculi of Identities and Inclusions. In {\it Recent Trends
   in Data Type Specification}, 
\bibitem[WM95b]{MW} S.~Meldal, M.~Walicki. A Complete Calculus for 
   Multialgebraic and Functional Semantics of Nondeterminism. 
   {\em ACM TOPLAS}, Vol.~17, No.~2, (1993).
\end{thebibliography} 

\end{document}

%%%%%%%%%% adt-bibliography
\begin{thebibliography}{MM99}
\bibitem[Bez90]{Bez} M.~Bezem. 
   Completeness of Resolution Revisited. 
   {\it Theoretical Computer Science}, 74, pp.27-237, (1990).
\bibitem[BG94a]{BG} L.~Bachmair, H.~Ganzinger. 
   Rewrite-Based Equational Theorem Proving with Selection and
   Simplification. {\it Journal of Logic and Computation}, 4(3),
   pp.~217--247 (1994).
\bibitem[BG94b]{BG249} L.~Bachmair, H.~Ganzinger. 
   Rewrite Techniques for Transitive Relations.
   {\it Proc. 9th IEEE Symposium on Logic in Computer Science}, pp.~384--393.
   IEE Computer Society Press (1994).
\bibitem[DJ90]{Der} N.~Dershowitz, J.-P.~Jouannaud. 
   Rewrite systems. In: J.~van Leeuwen (ed.)  {\it Handbook of
   theoretical computer science}, vol. B, chap. 6,
   pp.243-320. Amsterdam: Elsevier, (1990).
\bibitem[DM79]{DM} N.~Dershowitz, Z.~Manna. 
   Proving termination with multiset orderings. 
   {\it Communications of the ACM}, 22:8,pp.465-476, (1979).
\bibitem[DO92]{DO} A.~Dovier, E.~Omodeo, E.~Pontelli, G.-F.~Rossi. 
   Embedding finite sets in a logic programming language. 
   LNAI, vol.~660, pp.150-167, (1993).
\bibitem[Hes88]{PS1} W.H.~Hesselink. A Mathematical Approach to Nondeterminism
   in Data Types. {\it ACM Trans. on Programming Languages and Systems},
   10, pp.87-117, (1988).
\bibitem[Hus92]{Hus} H.~Hussmann. Nondeterministic algebraic
   specifications and nonconfluent term rewriting. {\it Journal of Logic
   Programming}, 12, pp.237-235, (1992).
\bibitem[Hus93]{HusB} H.~Hussmann. 
   {\it Nondeterminism in Algebraic Specifications and Algebraic Programs.}
   Birkh\"auser Boston, (1993).
\bibitem[Jay92]{Jay} B.~Jayaraman. Implementation of Subset-Equational 
   Programs. {\it Journal of Logic Programming}, 12:4, pp.299-324, (1992).
\bibitem[Kap88]{Kap} S.~Kaplan. Rewriting with a Nondeterministic Choice
   Operator. {\it Theoretical Computer Science}, 56:1, pp.37-57, (1988).
\bibitem[KW94]{KW} V.~Kriau\v ciukas, M.~Walicki.  Reasoning and Rewriting
   with Set-Relations I: Ground-Completeness.  In {\it Proceedings of
   CSL'94}, LNCS, vol.~933, pp.~264--278, (1995).
\bibitem[LA93]{LA} J.~Levy, J.~Agust\'i. Bi-rewriting, a term rewriting
   technique for monotonic order relations. In {\it RTA'93}, LNCS, 
   vol.~690, pp.17-31, (1993).
\bibitem[PP91]{PP} J.~Pais, G.E.~Peterson. Using Forcing to Prove Completeness
   of Resolution and Paramodulation. {\it Journal of Symbolic Computation}, 
   11:(1/2), pp.3-19, (1991).
%\bibitem [S-A92]{S-A} R.~Socher-Ambrosius. 
%   {\it Completeness of Resolution and Superposition Calculi}.
%   Technical Report
%   MPI-I-92-224, Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, (1992).
\bibitem [SD86]{SD} J.~Schwartz, R.~Dewar, E.~Schonberg, E.~Dubinsky. 
   {\it Programming with sets, an introduction to SETL. }
   Springer Verlag, New York, (1986).
%\bibitem[Sto93]{Sto} F.~Stolzenburg. 
%   {\it An Algorithm for General Set Unification.}
%   Workshop on Logic Programming with Sets, ICLP'93, (1993).
\bibitem[Wal93]{Mich} M.~Walicki. 
   {\it Algebraic Specifications of Nondeterminism.}
   Ph.D. thesis, Institute of Informatics, University of Bergen, (1993).
% Selected papers of joint 10th ADT and 5th COMPASS workshops, S.~Margherita, Italy, May 30 - June 3, 1994},
   LNCS, vol.~906, pp.~453--468 (1995).
\bibitem[WM95b]{MW} M.~Walicki, S.~Meldal. A Complete Calculus for 
   Multialgebraic and Functional Semantics of Nondeterminism. 
   {\it ACM Trans. on Programming Languages and Systems}, vol.~17, No.~2,
   pp.~366--393 (1995).
\end{thebibliography} 
