\documentclass[11pt]{letter}
\usepackage{euscript}
 \makeatletter
 \input{../a4wide}
 \makeatother

\begin{document}
%\date{30 Mai, 1996}
\begin{letter}{ {\bf ACM: CSUR}, Managing Editor\\
  1515 Broadway,\\
 New York,  NY 10036, USA }
\address{Micha{\l} Walicki \\ University of Bergen, Dept. of Informatics\\
         HiB, 5020 Bergen, NORWAY \\
         {\small{{\em e-mail:} michal@ii.uib..no}} }

%\signature{With best regards \\[4ex] Micha{\l} Walicki}

\opening{{\bf Re:} Proofs for 
{\em ``Algebraic Approaches to Nondeterminism -- an Overview''}}
Dear Editor,\\[1ex]
I have used some, perhaps, not quite standard convensions when marking the set of proofs:
\begin{itemize}
\item[{}] appears typically at places requiring correction;
\item[{}] appears at some points marked originally by your staff and indicates that 
 I do not understand why it was marked and, in any case, that I do not see anything wrong there;
\item[{}] marks 2-3 places where I'd like to add a single word/phrase not occurring in 
the original manuscript.
\item[{}] indicates suggested linebreak;\\
There are a lot of linebreaks, especially within mathematical formulae, which I find
very confusing. Unless they are motivated by some standard conventions, I would appreciate if
they could be changed according to my suggestions.
\item Some places I specified the \LaTeX-command for the symbol which was wrongly typeset.
%\end{itemize}
%Otherwise, a couple of points (the first two concerning changes made by your staff):
%\begin{itemize}
\item Some figures got red marks. \\
The change of some fonts (from {\sf{$\backslash$mathcal}} to some {\sf script} font) 
made by your staff was not reflected in the figures.
Unfortunately I was unable to find the same font you used. 
In the new set of figures, 
enclosed with this letter, the symbols (corresponding to the original 
ones $:{\cal I}$, ${\cal P}$, ${\cal C}$) were typeset using the font
 {\sf Euler Script} which I found closest to the one in the text (i.e., 
 $\EuScript{I}$,  $\EuScript{P}$, $\EuScript{C}$)
In the new set, the figures affected by this change (with red marks in the proofs) are marked 
``(fonts modified)''.
\item
Your staff replaced most (if not all) appearances of the word ``above'' by ``preceding''. 
A couple of places I felt that the result was unfortunate and suggested a new formulation.
\end{itemize}

\vspace*{6ex}
\hspace*{18em} With best regards \\[8ex]
\hspace*{19em} Micha{\l} Walicki

\end{letter}
\end{document}