\documentclass[10pt]{article}
%%\documentstyle[a4wide,10pt]{article}
\makeatletter
\input{a4wide}
\makeatother

%%%\setlength{\textwidth}{17cm}
%%%\setlength{\evensidemargin}{-.8cm}
%%%\setlength{\oddsidemargin}{-.8cm}

%% \makeatother
\input{defs}
\input xypic

\title{Cut-Elimination and Other Gimmicks\\ for Logic of Nondeterminism}

\author{ {\em Regimantas ~Pliu\v skevi\v cius}\ \thanks{Department of Mathematical Logic,
Vilnius Institute of Mathematics and Informatics, Lithuania,
{\sl \{logica@sedcs.mii2.lt\}}} \\
{\em Micha{\l}~Walicki}\ \thanks{Department of Informatics, University of Bergen, 5020
Bergen, Norway, {\sl \{michal@ii.uib.no\}}} \\
{\em Jurate~Sakalauskait\.e }$^*$ \\
{\em Aida~Pliu\v skevi\v cien\.e }$^*$ 
}

\begin{document}


\input{neqA}

\input{neqB}

\input{neqC}

\section{Related work.}\label{se:related}
The first, to our knowledge, reasoning system for multialgebras was introduced 
by Hu{\ss}mann in
\cite{Hus}. It dealt only with simple (atomic) inclusions and, more significantly, 
was not complete.
A complete extension of this calculus was given in \cite{Mich}.
The present system $NEQ_5$ is a significant improvement of the original
calculi from \cite{Mich} and \cite{Top}, particularly, due to cut-freeness.

Another cut-free, Gentzen-style calculus was introduced in \cite{BK95}.
It addresses proofs of tautologies (no specific axioms) and so it is a counterpart
of our $NEQ_6$.
% -- $NEQ_5$ is essentially an extension of this system.
\begin{itemize}\MyLPar
\item 
The main difference is that \cite{BK95} address partial multialgebras, i.e. ones where
some terms may denote empty set. Our calculi work only for total algebras. 
Nevertheless, it seems that partiality is captured by ...
\item
There are two differences in the language. The first one concerns
the special ``let'' term-construction: `$let\ x:=t\ in\ f(x,x)$' is a possible term. 
Although absent from our language, it is
expressible, except for the cases when a term with $let$ occurs in the RHS
of inclusion. 
\item 
On the other hand, our system
deals not only with inclusion, but also with the determinisitc equality which
is not a part of the language in \cite{BK95}.
Thus, our results show that the
only rules needed for their case are the first axiom and the inclusion rules
from figure \ref{fi:neq6}.
\item
The proofs in the calculus from \cite{BK95} proceed essentially in a bottom-up 
fashion by 1) replacing (sub)terms of concern by variables and introducing 
additional assumptions about the relations between them, and then 2) performing 
the deductions using rules handling variables. In general, this leads to complex
proof-{\em trees} -- in $NEQ_6$ all the proofs are linear.
\item
The restrictions on the inclusion and elimination rules are essentially the same
in both systems.
However, we have
made it explicit that the elimination rules (which are present in \cite{BK95}, 
even though no specific axioms are admitted)
are needed only in the presence of 
specific axioms -- when no such axioms are involved they can be eliminated and $NEQ_6$
is a complete system for deriving all multialgebraic tautologies.
\end{itemize}
The sufficiency of $\elNEQ$ is a significant strengthening of these earlier results.
In addition, the cut-freeness of the system from \cite{BK95} corresponds to the 
situations when
no specific axioms are present (i.e., cut-freeness of $\elNEQ$). 
We have shown how such axioms can be treated explicitly by
the more specific cut rules instead of general cut. Finally, other rules of the 
calculus have
been given very specific restrictions. All these results enhance significantly 
proof search.

\subsection*{Acknowledgment}
We thank Valentinas Kriau\v ciukas and Sigurd
Meldal for the comments and help they provided during our work on this paper.

The first three authors were supported by a grant from the Norwegian Research Council.

\begin{thebibliography}{MM99}\MyLPar
\bibitem[BK95]{BK95} M.~Bia{\l}asik, B.~Konikowska. 
   {\em Reasoning with Nondeterministic Specifications,} Tech.~Rep. no.~793, IPI PAN, Warszawa,
   (1995).
\bibitem[Hus90]{Hus} H. Hu{\ss}mann, {\em Nondeterministic Algebraic Specifications},
Ph.D. thesis, Fakult\"{a}t f\"{u}r Mathematik und Informatik, Universit\"{a}t Passau, 
 (1990).
\bibitem[Kan63]{K} S.~Kanger,
   {\em A simplified proof method for elementary logic,}
   in Comput. Progr. and Formal Systems, North-Holland, Amsterdam, (1963).
\bibitem[Pli71]{Aida1} A.~Pliu\v skevi\v cien\.e, 
   {\em Elimination of cut-type rules in axiomatic theories with equality,}
   Seminars in mathematics V.A.Steklov Mathem. Inst., Leningrad, 16, 90-94, (1971).
\bibitem[Pli73]{Aida2} A.~Pliu\v skevi\v cien\.e,
   {\em Specialization of the use of axioms for deduction search in 
   axiomatic theories with equality,}
   J. Soviet Math., 1, (1973).
\bibitem[PWM94]{LFCS} A.~Pliu\v skevi\v cien\.e, R.~Pliu\v skevi\v cius,
   M.~Walicki, S.~Meldal,
   {\em On specialization of derivations in axiomatic equality theories,}
   in Proc. of LFCS'94, LNCS vol. 813, (1994).
\bibitem[Wal93]{Mich} M.~Walicki,
   {\em Algebraic Specifications of Nondeterminism,}
   Ph.D. thesis, Institute of Informatics, University of Bergen, (1993).
\bibitem[WB97]{WB} M.~Walicki, M.~Bia{\l}asik, {\em Relations, Multialgebras and 
   Homomorphisms}, Tech.Rep., no~838, Polish Academy of Sciences, Institute of
  CS, (1997).
\bibitem[WB95]{Broy} M.~Walicki, M.~Broy, Structured Specifications and Implementation of
   Nondeterministic Data Types, {\em Noric Journal of Computing}, vol.~2, (1995).
\bibitem[WM95a]{WM} M.~Walicki, S.~Meldal, 
   Multialgebras, Power Algebras and Complete Calculi of Identities and 
       Inclusions, {\em Recent Trends in Data Type  Specification}, LNCS, vol. 906 (1995).
\bibitem[WM95b]{Top} M.~Walicki, S.~Meldal, 
   A Complete Calculus for the Multialgebraic and Functional Semantics of Nondeterminism,  
       {\em ACM ToPLaS}, vol.~17, no.~2, (1995).
\end{thebibliography} 


\end{document}
